---
title: "Regression Analysis: Pitcher Performance"
author: "Ricky Trujillo"
date: "10/14/2020"
output: 
  pdf_document:
    extra_dependencies: ["amsmath", "amssymb", "latexsym","tabularx"]
    toc: true
fontsize: 12pt
geometry: "left=1cm,right=1cm,top=1cm,bottom=1cm"
documentclass: article
latex_engine: pdflatex
---

```{r setup, include = FALSE, warning=FALSE}
require(devtools)
require(neuralnet)
install_github("BillPetti/baseballr")
library(baseballr)
library(leaps)
library(olsrr)
library(ggcorrplot)
library(randomForest)
library(pls)
library("class")
```

\newpage
# Abstract
This article evaluates pitchers' pitching performance during the 2008 to 2019 Major League Baseball(MLB) seasons. A pitcher's Estimated Run Average (ERA) is crucially important to look at when assessing pitching performance as a lower ERA indicates a better assessment of the given pitcher. That being the case I will be taking a look at which variables from the given dataset causes a pitcher's ERA to fluctuate for better or worse. In doing so, I am immediately assuming non-linearity in the dataset and applying Random Forest Regression, Spline Regression, and Generalized Additive Models to study and analyze the relationships the other predictors have to the ERA response. 

\newpage

# Baseball-Reference Data 

## Baseballr package
" baseballr is a package written for R focused on baseball analysis. It includes functions for scraping various data from websites, such as FanGraphs.com, Baseball-Reference.com, and baseballsavant.com. It also includes functions for calculating metrics, such as wOBA, FIP, and team-level consistency over custom time frames."

*The latest package, baseballr 0.6.0, published by Bill Petti had its latest version release on January 07, 2020
*[documentation](http://billpetti.github.io/baseballr/about/) 

## Variable Glossary 
$$\boldsymbol{bbref\_id}: \text{ Baseball Reference ID for Players } $$
$$\boldsymbol{season}:\text{Respective year to each season time period}$$
$$\boldsymbol{Name}: \text{Name of Pitcher} $$
$$\boldsymbol{Age}: \text{Age of Pitcher}$$
$$\boldsymbol{Level}:\text{League of Pitcher (i.e. MLB National League and/or MLB American League)}$$
$$\boldsymbol{Team}:\text{MLB Team of Pitcher}$$
$$\boldsymbol{G}:\text{Number of Games Pitched}$$
$$\boldsymbol{GS}:\text{Grand Slam}$$
$$\boldsymbol{W}:\text{Number of games won}$$
$$\boldsymbol{L}:\text{Number of games lost }$$
$$\boldsymbol{SV}:\text{Save}$$
$$\boldsymbol{IP}:\text{Number of Innings Pitched}$$
$$\boldsymbol{H}:\text{Number of Hits}$$
$$\boldsymbol{R}:\text{Number of Runs}$$
$$\boldsymbol{ER}:\text{Earned Run}$$
$$\boldsymbol{uBB}:\text{unintentional base on ball}$$
$$\boldsymbol{BB}:\text{Base on balls (or walks)}$$
$$\boldsymbol{SO}:\text{Strikeouts }$$
$$\boldsymbol{HR}:\text{Number of Homeruns}$$
$$\boldsymbol{HBP}:\text{Hit-by-pitch}$$
$$\boldsymbol{ERA}:\text{Earned Run Average}$$
$$\boldsymbol{AB}:\text{At-Bat}$$
$$\boldsymbol{X1B}:\text{Number of Singles}$$
$$\boldsymbol{X2B}:\text{Number of Doubles}$$
$$\boldsymbol{X3B}:\text{Number of Triples}$$
$$\boldsymbol{IBB}:\text{Intentional Base on Ball}$$
$$\boldsymbol{GDP}:\text{Ground into Double Play}$$
$$\boldsymbol{SF}:\text{Sacrifice Fly}$$
$$\boldsymbol{SB}:\text{Stolen Base}$$
$$\boldsymbol{CS}:\text{Caught Stealing}$$
$$\boldsymbol{PO}:\text{Putout}$$
$$\boldsymbol{BF}:\text{Batters Faced}$$
$$\boldsymbol{Pit}:\text{Number of Pitches}$$
$$\boldsymbol{Str}:\text{Number of Strikes}$$
$$\boldsymbol{StL}:\text{Strikes per Loss}$$
$$\boldsymbol{StS}:\text{Strikes per Save}$$
$$\boldsymbol{GB.FB}:\text{Ground Ball/ Fly Ball}$$
$$\boldsymbol{LD}:\text{Line Drives}$$
$$\boldsymbol{PU}:\text{Popup }$$
$$\boldsymbol{WHIP}:\text{Walk and Hits per Innings Pitched}$$
$$\boldsymbol{BAbip}:\text{Batting Average on balls in play}$$
$$\boldsymbol{SO9}:\text{Average Strikeout over 9 innings}$$
$$\textbf{SO.W}:\text{Strikeouts before another Walk}$$
$$\boldsymbol{SO\_perc}: \text{Strikeout Percentage}$$
$$\boldsymbol{uBB\_perc}: \text{Unintentional Base on Ball Percentage}$$
$$\boldsymbol{SO\_uBB}: \text{Strike outs to normal walks ratio}$$


## Baseballr Data Setup 
What to Discuss:
  - Mention the season years in accordance to the dates of each season
  - NA observations were omitted which led to a size reduction 
  - combined all into one dataframe called Pitchers
  
```{r, include = FALSE, warning=FALSE}
pitcher_2008= daily_pitcher_bref(t1 = "2008-03-25", t2 = "2008-10-29")
pitcher_2009= daily_pitcher_bref(t1 = "2009-04-05", t2 = "2009-11-04")
pitcher_2010 = daily_pitcher_bref(t1 = "2010-04-04", t2 = "2010-11-01")
pitcher_2011 = daily_pitcher_bref(t1 = "2011-03-31", t2 = "2011-09-28")
pitcher_2012 = daily_pitcher_bref(t1 = "2012-03-28", t2 = "2012-10-28")
pitcher_2013 = daily_pitcher_bref(t1 = "2013-03-31", t2 = "2013-10-30")
pitcher_2014 = daily_pitcher_bref(t1 = "2014-03-31", t2 = "2014-10-29")
pitcher_2015= daily_pitcher_bref(t1 = "2015-04-05", t2 = "2015-11-01")
pitcher_2016= daily_pitcher_bref(t1 = "2016-04-03", t2 = "2016-10-02")
pitcher_2017= daily_pitcher_bref(t1 = "2017-04-03", t2 = "2017-11-01")
pitcher_2018= daily_pitcher_bref(t1 = "2018-03-29", t2 = "2018-10-28")
pitcher_2019= daily_pitcher_bref(t1 = "2019-03-28", t2 = "2019-09-29")
Pitchers = na.omit(rbind(pitcher_2008,pitcher_2009,pitcher_2010,pitcher_2011,pitcher_2012,pitcher_2013,pitcher_2014, pitcher_2015,pitcher_2016, pitcher_2017, pitcher_2018, pitcher_2019))
``` 

What to Discuss:
  - The different variable types and focus in on Level being a factor variable out of 4
      - Explain how Level is of only interest from the factor variables
      - Leads into contrast coding 
```{r}
str(Pitchers)
```

Contrast Coding: 
```{r}
Pitchers$Level = factor(Pitchers$Level)

for (i in c(1:46))
    if (class(Pitchers[,i]) == "factor")
        print(levels(Pitchers[,i]))

set.seed(1)
response = Pitchers$ERA
predictors = Pitchers[ ,-c(1,3,6)]

contrastencode <- function(z,basename){
    # z = factor variable
    # basename = names of the variable
    nl = length(levels(z))
    newmat = matrix(0,nrow=length(z),ncol=(nl-1)) 
    con = contrasts(z)
    for (j in c(1:ncol(con))){ 
        thislevel = colnames(con)[j]
        theserows = which(z == thislevel) 
        newmat[theserows, j] = 1 
    }
    colnames(newmat) = paste(basename,".is.",colnames(con),sep='')
    return(newmat)
}


contrast_vars = list(NULL)
contrast_vars[[1]] = contrastencode(predictors[,3],names(predictors)[3])
newPitchers = data.frame(ERA=response)
newPitchers = cbind(newPitchers, contrast_vars[[1]], predictors[ ,-3])
rownames(newPitchers) <- 1:nrow(newPitchers)
```
What to discuss: 
  - Level is a factor variable with 4 levels (list those levels)
  - change its qualitative stance into a quantitative one to see if the League of a given pitcher reflects on their ERA score 
  - Describe the function (Credit Prof. Bhat)
  - newPitchers is a new dataframe consisted of the different levels from constrast coding as dummy variables so to say
      - describe the L-1 aspect of it being included in it already (MLB-NL = 0 0 0)

# Exploratory Data Analysis

Take a look at the summary logistics
    - Thought to include it but I was hoping to get more insight on what values or metrics of this I should look at 
            - Looking to get feedback here 
```{r}
summary(newPitchers)
```

```{r}
plot(newPitchers$season, newPitchers$ERA, 
     xlab="Season", 
     ylab="ERA", 
     main="Plots of ERA from 2008-2019",
     sub="Figure 1: Plotting each player's ERA for each season ranging from 2008 to 2019")
```
What to describe:
- Describe how it is plots across all the years of interest 
- The averages of those ERA values 
- The outliers (a few) 


```{r}
ggcorrplot(cor(newPitchers[,-c(1,5)]))
```
What to describe:
- Which variables are negatively correlated and positively correlated with ERA


```{r}
set.seed(1)
names(newPitchers) <- make.names(names(newPitchers))
split = sample(2, nrow(newPitchers), replace=TRUE, prob=c(0.7,0.3))
train= newPitchers[split==1, ]
test=newPitchers[split==2, ]
```
What to explain
  - Train-Test Split
  - 70%, 30% split
  - set seed to have reproducibility
  
\newpage



# PCR and PLA Regression
"DESCRIBE WHAT PCR IS"
  - Formulas
  - Parameters
```{r}
Pitcher_PCA =pcr(ERA~. - Age - season, data=train, validation="CV")
summary(Pitcher_PCA)
validationplot(Pitcher_PCA,val.type="MSEP")
validationplot(Pitcher_PCA,val.type="R2")
coefplot(Pitcher_PCA)
predplot(Pitcher_PCA)

Pitcher_PCApred = predict(Pitcher_PCA, test,scale=TRUE, ncomp=10)
mean((Pitcher_PCApred-test$ERA)^2)*100

Pitcher_PCApred2 = predict(Pitcher_PCA, test, scale=TRUE, ncomp=30)
mean((Pitcher_PCApred2-test$ERA)^2)*100

```
What to describe: 
  - The number of components to where we get 90% percent of variance explained
  - How many components lead to an appropriate low value without having to use all variables
      - otherwise if we have all the variables, we do nothing in reducing the dimensionality 
  - describe the MSEP plot trend
  - describe the coefficient plot 
  - describe the prediction plot
  - describe the model prediction when its 10 components and when its 30 components
  

"DESCRIBE WHAT PLSR IS" 
  - Formulas
  - Parameters
```{r}
Pitcher_PLS =plsr(ERA~. - Age - season, data=train, validation="CV")
summary(Pitcher_PLS) 
validationplot(Pitcher_PLS,val.type="MSEP")
validationplot(Pitcher_PLS,val.type="R2")
coefplot(Pitcher_PLS)
predplot(Pitcher_PLS)
```
- The number of components to where we get 90% percent of variance explained
  - How many components lead to an appropriate low value without having to use all variables
      - otherwise if we have all the variables, we do nothing in reducing the dimensionality 
  - describe the MSEP plot trend
  - describe the coefficient plot 
  - describe the prediction plot
\newpage

# Neural Networks
~ Still needs more work 
  - Formulas
  - Parameters
```{r}
Pitchers_NN=neuralnet(ERA~. - Age - season,data=train, hidden=c(5,3), rep=3, act.fct = "logistic",linear.output = FALSE) 
plot(Pitchers_NN)
```
- Describe the effects of hidden layers (i.e. Depth vs Breadth) and how it can improve the model or not
    - Illustrate with a figure and subtitle the idea of how the Neural Network is doing
    - Need to figure out how to use it to predict the test observations??
    
\newpage

# Random Forest Regression
First Case: 
```{r}
Pitchers_RFM = randomForest(ERA~. - Age - season, data=train, mtry=20, importance=TRUE)
Pitchers_RFM
ERA_pred = predict(Pitchers_RFM, test)
cbind(test$ERA, ERA_pred)
plot(test$ERA, ERA_pred)
abline(0,1)
mean((test$ERA-ERA_pred)^2)*100
```
What to discuss: (20 variables but 500 trees)
  - Number of variables tried at each split
  - Number of trees
  - Variance Explained
  - Take a look at the actual values and predicted values side by side in a table
  - see how they compare in a plot, then plot a linear line through it and calculate the MSE

  
```{r}
Pitchers_RFM2 = randomForest(ERA~. - Age - season, data=train, mtry=20, ntrees=100, importance=TRUE)
Pitchers_RFM2
ERA_pred2 = predict(Pitchers_RFM2, test)
plot(test$ERA, ERA_pred2)
abline(0,1)
mean((test$ERA-ERA_pred2)^2)*100
```
What to discuss:(20 variables but 300 teress ) 
  - Number of variables tried at each split
  - Number of trees
  - Variance Explained
  - Take a look at the actual values and predicted values side by side in a table
  - see how they compare in a plot, then plot a linear line through it and calculate the MSE
  
  
```{r}
Pitchers_RFM3 = randomForest(ERA~. - Age - season, data=train, mtry=10, importance=TRUE)
Pitchers_RFM3
ERA_pred3 = predict(Pitchers_RFM3, test)
cbind(test$ERA, ERA_pred3)
plot(test$ERA, ERA_pred3)
abline(0,1)
mean((test$ERA-ERA_pred3)^2)*100
```
What to discuss: (10 variables, but 500 trees)
  - Number of variables tried at each split
  - Number of trees
  - Variance Explained
  - Take a look at the actual values and predicted values side by side in a table
  - see how they compare in a plot, then plot a linear line through it and calculate the MSE


```{r}
Pitchers_RFM4 = randomForest(ERA~. - Age - season, data=train, mtry=10, ntrees=100, importance=TRUE)
Pitchers_RFM4
ERA_pred4 = predict(Pitchers_RFM4, test)
plot(test$ERA, ERA_pred4)
abline(0,1)
mean((test$ERA-ERA_pred4)^2)*100
```
What to discuss: (10 variables, but 100 trees)
  - Number of variables tried at each split
  - Number of trees
  - Variance Explained
  - Take a look at the actual values and predicted values side by side in a table
  - see how they compare in a plot, then plot a linear line through it and calculate the MSE
  
  
```{r}  
#THIS IS GOING TO BE DONE FOR ALL DIFFERENT VARIATIONS OF THE RFM ABOVE
# Right now it is only based on the first one
importance(Pitchers_RFM)
varImpPlot(Pitchers_RFM)
```
What to Discuss: 
  - The largest values are the most significant
      - compare these to what the correlation depicted before
      
\newpage


# References
Petti, Bill. “Baseballr.” Root Mean Squared Musings, 2020, billpetti.github.io/baseballr/. 


